---
title: "README - Customer Churn Models"
author: "Martin Frigaard"
date: "current version: `r Sys.Date()`"
output: github_document
---


```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
require(tidyverse)
library(plotly)
require(janitor)
require(skimr)
library(mosaic)
library(inspectdf)
library(visdat)
library(DT)
# base options ----
base::options(
  tibble.print_max = 25,
  tibble.width = 78,
  scipen = 100000000,
  max.print = 999999
)
# knitr chunk options ----
knitr::opts_chunk$set(
  echo = TRUE, # show/hide all code
  # results = "hide", # hide results
  tidy = FALSE, # cleaner code printing
  comment = "#> ", # better console printing
  eval = TRUE, # turn this to FALSE stop code chunks from running
  message = TRUE, # show messages
  fig.width = 7, # figure width
  fig.height = 5, # figure height
  warning = FALSE, # show warnings
  size = "small", # size of the text
  fig.path = "figs/"
) # location of files
# knitr knit settings ----
knitr::opts_knit$set(
  width = 78
)
```

# Motivation

This project is for [Demystify Data Science at EscapeSF](https://www.bigsquid.com/san-francisco-machine-learning-event-2019).

> Business leaders need to make decisions at the speed of business, and yet traditional machine learning projects can take months before they ever see the light of day. 

> We want you to join us for an afternoon of drinks, food, and an escape room with Big Squid machine learning experts and analytics powerhouses from Paradigm Data. You’ll hear what it takes to prepare for machine learning initiatives, how to efficiently develop and deploy ML solutions, and how to democratize data science through your organization.

The documentation for this data set is available [here](https://www.kaggle.com/blastchar/telco-customer-churn) and in the readme file that comes with the Kaggle dataset. 

## Set the stage: what is the problem?

What decision needs to be made? What problem needs to be figured out? What numbers need to be calculated? This is the 'main thing' (and the main thing is to keep the main thing the main thing).

## Know the characters: what are the data?

What has been measured? What needs to be measured? What can we do with the information we have? What new information will we need when we've used everything we have?

```{r data-dict}
fs::dir_tree("data/raw", regex = "dict")
```

## Connect the business problem to a measurable objective

The context statement for the data in this example is, "*Predict behavior to retain customers. You can analyze all relevant customer data and develop focused customer retention programs.*"

> "Customer churn, when a customer ends their relationship with a business, is one of the most basic factors in determining the revenue of a business. You need to know which of your customers are loyal and which are at risk of churning, and you need to know the factors that affect these decisions from a customer perspective. This code pattern explains how to build a machine learning model and use it to predict whether a customer is at risk of churning. This is a full data science project, and you can use your model findings for prescriptive analysis later or for targeted marketing."

Our measurable objective with this data set is: **Identify predictors for customer churn**.

## Map the measurable objective to the data 

The hardest part of using data to solve problems is asking questions in a way that a computer can answer them. We need to ask the question in a way where we can leverage the power of computation--this means fitting our objectives into an analytical framework capable of answering our questions.

### Clearly defining the question/problem of interest

1. We can tell that customer retention (or it's inverse, churn) is the *outcome*. What are the behavioral variables?  

2. Were all the data collected/measured at the same time? If not, is the time different significant? 

Is the customer data current? Was it defined in the same way at each time point? For example, if a customer indicated they were not a senior citizen (i.e. `SeniorCitizen` is `0`) at the beginning of their tenure, but after 60 months they've retired and are now living on a fixed income, this data has changed and no longer captures the measurement relevant to our business question (i.e. *What behaviors drive customer churn?*).

3. What is the level of granularity in this data set? 

 - Is arranged by each customer's monthly bill, so that each `customerID` is repeated 12 times per year, with the services listed for each month, or
 
 - Is this an aggregated/summarized data set with a single measure for the monthly or quarterly bill?

For example, assume a customer has been receiving services for 45 months. Can we tell if the number of services for that customer has changed over the course of their `tenure`? *No.* However, having the data structured in it's current format, we can use `tenure` as a single attribute to predict `Churn`. That would be more complicated to analyze and interpret if we had monthly data.

These questions are also important when we're designing experiments based on the findings. For example, if the business tolerance is "impact within 30 days", then we need to have the data collected and analyzed at the corresponding level. 

Now that we have some questions about what these data look like, we can start getting some answers. Business questions come from a wide audience; stakeholders, revenue, competition, etc., but the analytic framework and available metrics will ultimately frame these questions into what can be used to drive decisions. 

We will import these data and see their structure. 

## The process

Every data project has the same steps, just at different scales. They are: 

1. Import the data into some software tool  
2. Prepare (wrangle, clean, process, etc.) the data for visualization or analysis  
3. Explore the data with summary statistics, graphs, figures, etc.  
4. Model the data (use it to make predictions, forecast, etc.)  
5. Communicate the results to the necessary decision makers  

In this talk, we are going to focus on steps 1 - 3. 

## Importing the data

This isn't a process to get hung up on, but it's really just about finding the right tool for the job. SQL is a great tool for keeping track of your customers, products, etc., but it can't visualize or model. This isn't a problem until you want to visualize or model (which will be sooner than you think). 

The tools you use and train your staff will ultimately dictate how they look at your data, and the kinds of solutions they'll create ("if the only tool you have is a hammer, to treat everything as if it were a nail" - [Law of instrument](https://en.wikipedia.org/wiki/Law_of_the_instrument)).

```{r import-WA_Fn-UseC_-Telco-Customer-Churn}
# fs::dir_ls("data/raw")
TelcoDataRaw <- readr::read_csv("data/raw/WA_Fn-UseC_-Telco-Customer-Churn.csv")
TelcoDataRaw %>% count(SeniorCitizen)
```

Standardize the variable names with `janitor::clean_names()`

```{r clean-names-TelcoData}
TelcoData <- TelcoDataRaw %>% janitor::clean_names(case = "snake")
TelcoData %>% dplyr::glimpse(78)
```


The first question we should ask is what the variables (columns) in the data represent. We can break these columns down into the following four categories:

### Things we know about the customer

`customer_id` - unique identifier

`gender` - male or female

`senior_citizen` - Are they a senior citizen?

`partner` - Do they have a partner?

`dependents` - Do they have any dependents?

```{r customer-data, message=FALSE, warning=FALSE}
TelcoData %>% 
  dplyr::select(customer_id:dependents) %>% 
  skimr::skim_to_wide()
```

```{r missing-customer-data}
# missing data 
TelcoData %>% 
  dplyr::select(customer_id:dependents) %>% 
  inspectdf::inspect_na(df1 = .) %>% 
  inspectdf::show_plot()
```

```{r numeric-customer-data-plot}
TelcoData %>% 
  dplyr::select(customer_id:dependents) %>% 
  inspectdf::inspect_num(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

```{r categorical-customer-data-plot-02}
TelcoData %>% 
  dplyr::select(customer_id:dependents) %>% 
  inspectdf::inspect_cat(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

This shows us that the senior citizen variable is coded as `0` and `1`, but should also have a categorical equivalent.

```{r senior_citizen_num}
# call this _num
TelcoData <- TelcoData %>% 
  dplyr::rename(senior_citizen_num = senior_citizen)

# create categorical variable
TelcoData <- TelcoData %>% 
  dplyr::mutate(senior_citizen = case_when(
    senior_citizen_num == 1 ~ "Yes",
    senior_citizen_num == 0 ~ "No")) 
# put this in right place
TelcoData <- TelcoData %>% 
    dplyr::select(customer_id:gender,
                  dplyr::starts_with("senior"),
                  partner:churn) 
```

Check this just to verify.

```{r check-senior_citizen}
# check the new variable
TelcoData %>% 
  dplyr::count(senior_citizen_num, senior_citizen) %>% 
  tidyr::spread(senior_citizen, n)
```

Ok we're looking better.

```{r recheck-senior_citizen}
TelcoData %>% 
  dplyr::select(customer_id:dependents) %>% 
  inspectdf::inspect_cat(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

Now we see very few of them are senior citizens. 

### Variables on the customer's relationship to our products & services:

Our customer's relationship to our products/services is captured in the nine items below in this data set. 

`phone_service` - Did they purchase phone service?

`multiple_lines` - Did they purchase multiple lines?

`internet_service` - Did they purchase internet services?

`online_security` - Did they purchase online security

`online_backup` - Did they purchase online backup?

`device_protection` - Did they purchase device protection?

`tech_support` - Did they purchase tech support?

`streaming_tv` - Did they purchase streaming TV?

`streaming_movies` - Did they purchase streaming movies?

The `tenure` column is capturing the length of time each customer has had a relationship with our products/services. 

#### Descriptive statistics of the product/service variables

The `skimr::skim()` gives us a breakdown of the counts, missing, min, max, mean, sd, and other relevant statistics on this data frame. 

```{r skim-product-data, message=FALSE, warning=FALSE}
TelcoData %>%
  dplyr::select(tenure:streaming_movies) %>%
  skimr::skim()
```

#### Missing product/service data

```{r missing-product-data}
# missing data 
TelcoData %>% 
  dplyr::select(tenure:streaming_movies) %>% 
  inspectdf::inspect_na(df1 = .) %>% 
  inspectdf::show_plot()
```

#### Numerical product/service data

`tenure` - how long have they been a customer?

```{r numeric-product-data}
TelcoData %>% 
  dplyr::select(tenure:streaming_movies) %>% 
  inspectdf::inspect_num(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

#### Categorical product/service variables 

These are the nine categorical variables from the services we provide. 

```{r categorical-product-data}
TelcoData %>% 
  dplyr::select(tenure:streaming_movies) %>% 
  inspectdf::inspect_cat(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```



### Things we know about the customer's :

These columns tell us about how the customer has been paying us.

```{r billing-data, message=FALSE, warning=FALSE}
TelcoData %>% 
  dplyr::select(contract:total_charges) %>% 
  skimr::skim()
```

Missing billing data

```{r missing-billing-data, message=FALSE, warning=FALSE}
TelcoData %>% 
  dplyr::select(contract:total_charges) %>% 
  inspectdf::inspect_na(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

### Drop missing variables 

We should remove the missing variables from the `total_charges` data frame. 

```{r drop-na}
TelcoData <- TelcoData %>% tidyr::drop_na()

# recheck
TelcoData %>% 
  dplyr::select(contract:total_charges) %>% 
  inspectdf::inspect_na(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

Numeric billing data

```{r numeric-billing-data, message=FALSE, warning=FALSE}
TelcoData %>% 
  dplyr::select(contract:total_charges) %>% 
  inspectdf::inspect_num(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

Categorical missing data from the billing variables. 

```{r categorical-billing-data, message=FALSE, warning=FALSE}
TelcoData %>% 
  dplyr::select(contract:total_charges) %>% 
  inspectdf::inspect_cat(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

We can see the three most common categories for these variables. 

### These are things we can't directly effect

These are variables that we can't change easily, *and* changing them wouldn't reliably result in adding value to our existing customers. We want to keep our eyes on minimal changes we can implement that result in adding value to our existing customers. 

`contract` - What is their contract?

`paperless_billing` - Do they have paperless billing?

`payment_method` - What is their payment method?

```{r skim-payment-data}
TelcoData %>% 
  dplyr::select(contract:payment_method) %>% 
  skimr::skim_to_wide()
```

Below are the missing data from the payment variables.

```{r missing-payment-data}
# missing data 
TelcoData %>% 
  dplyr::select(contract:payment_method) %>% 
  inspectdf::inspect_na(df1 = .) %>% 
  inspectdf::show_plot()
```

These are the categorical variables in the data frame.

```{r categorical-payment-data}
TelcoData %>% 
  dplyr::select(contract:payment_method) %>% 
  inspectdf::inspect_cat(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

### The thing we are trying to predict:

The variables we want to change the dial on (i.e. outcomes) are the metrics attached to revenue. **Always think of outcomes in terms of dollars**. 

`monthly_charges` - What are the monthly charges?

`total_charges` - What are the total charges up to this point?

`churn` - Did they customer stay/leave?

```{r skim-outcome-data}
TelcoData %>% 
  dplyr::select(monthly_charges:churn) %>% 
  skimr::skim_to_wide()
```

Below are the missing data from the 

```{r missing-customer-data-plot}
# missing data 
TelcoData %>% 
  dplyr::select(monthly_charges:churn) %>% 
  inspectdf::inspect_na(df1 = .) %>% 
  inspectdf::show_plot()
```

```{r numeric-customer-data-plot-02}
TelcoData %>% 
  dplyr::select(monthly_charges:churn) %>% 
  inspectdf::inspect_num(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

```{r categorical-customer-data-plot}
TelcoData %>% 
  dplyr::select(monthly_charges:churn) %>% 
  inspectdf::inspect_cat(df1 = .) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```


Here we can see the `churn` is a `character` variable, and indicates if the customer stayed (`Yes`) or left (`No`). Before we can make any calculations with churn, we'll need to convert it to a numerical variable. 

```{r churn_num}
TelcoData <- TelcoData %>% 
  dplyr::mutate(churn_num = dplyr::case_when(
    churn == "Yes" ~ 1, 
    churn == "No" ~ 0))
# check
TelcoData %>% 
  count(churn, churn_num) %>% 
  spread(churn, n)
```

We can use this new `churn_num` variable to capture how much each category costs in terms of dollar amounts. 

### Data inspection

Now we want to get a feel for how much data we have on each variable. Examining the data as a whole will give us an overview of the complete/missing data in each of the categories we outlined above.

We can see tenure is largest below 20 months (the giant red bar). This tells us something. We should ask ourselves, "*how can I capture this relationship in the data frame?*"

We can get the total percent and number of customers who churned.

```{r TotalChurn}
# churn rate total format
TotalChurn <- TelcoData %>%
  dplyr::group_by(churn_num) %>%
  dplyr::summarize(`total churn` = n()) %>%
  dplyr::mutate(`churn percent` = `total churn` / sum(`total churn`))
TotalChurn
```


```{r Churn_X_Tenure}
ChurnXTenure <- TelcoDataRaw %>%
  # find all customers who churned
  filter(Churn == "Yes") %>% 
  # get a count of each number of months a customer has been with our company
  group_by(tenure) %>% 
  # add this number to the data set
  mutate(no_tenure = n()) %>%
  # 
  arrange(desc(no_tenure))
ChurnXTenure %>% head()
```


Below we can see the distribution of the numerical variables in the `TelcoDataRaw` dataset. 

```{r inspectdf_num}
inspectdf::inspect_num(df1 = TelcoDataRaw) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```

Here we see the `SeniorCitizen` variable is coded numerically (`0` = No, `1` = Yes), so this should be reformatted and displayed in the plot above.

## Wrangle 

We know we have two variables that need to be reformatted (`customerID` and `SeniorCitizen`). We will address this in the code below


```{r TelcoData}
TelcoData <- TelcoDataRaw %>% 
  dplyr::mutate(customerID = as.factor(customerID), 
                       SeniorCitizen = case_when(
                         SeniorCitizen == 1 ~ "Yes", 
                         SeniorCitizen == 0 ~ "No",
                         TRUE ~ NA_character_))
# re-graph
inspectdf::inspect_cat(df1 = TelcoData) %>% 
  inspectdf::show_plot(text_labels = TRUE) 
```

Now we can recheck the numerical variables. 

```{r recheck-numerical}
inspectdf::inspect_num(df1 = TelcoData) %>% 
  inspectdf::show_plot(text_labels = TRUE)
```


## Tidy

Do the number of rows equal the number of unique customers? Or are there duplicates? This code tells us if the number of unique `customerID`s is identical to the number of rows in `TelcoData`.

```{r check-unique}
base::identical(x = TelcoData %>% dplyr::distinct(customerID) %>% nrow(),
                y = TelcoData %>% nrow())
```

Good. This helps us understand what we have in terms of data per customer. 

### Get a case study

Identify a case to represent a customer's experience in your dataset.

```{r 7795-CFOCW}
TelcoCaseStudy <- TelcoDataRaw %>% 
  dplyr::sample_n(size = 1)
TelcoCaseStudy
```

A single case study gives us a picture of how a customer moves through the data set. 

## Know your data

"The main thing is to keep the main thing the main thing." -> Nick Saban

In any dataset, there are probably more variables than you'll need to address the question. It's important to think about 1) *what did we measure?* (i.e. unit of analysis?), 2) *who/what did we measure it on?* (what attributes do we have on this unit of measurement?), 3) *what attributes can we act on?* (is there anything in this dataset that can effect and change the outcome?), 4) *what decisions will the new information effect?* (e.g. will we run experiments, add/remove features to our product, change marketing program, etc.), 5) and *who will be using this information to make a decision?* (is this going to my CFO, marketing manager, software engineer, etc.)

Here we have a case study, and this allows us to start telling the story of what we're seeing. The case study is the character in this story.

1. This customer is a [`gender`], is/isn't a senior citizen [`SeniorCitizen`], has/doesn't have a partner [`Partner`], has/doesn't have dependents [`Dependents`] 

2. They've paying for our products/services for 45 months   
3. They have four services (DSL, online security, device protection, and tech support)   

We know this case is an existing customer (`Churn` = `No`), and that he's paid `42.3` per month. 

#### Know how things are measured

All of these questions are based on binary (`Yes`/`No`) responses to the following services.

### Solve for case 

Tidy these data for the case. 

```{r}
TelcoCaseStudy %>% 
  tidyr::gather(key = "service", 
                value = "purchase_status", 
                PhoneService:StreamingMovies) %>% 
  count(purchase_status)
```

The `MonthlyCharges` and `TotalCharges` are likely summarized before they ended up in this data set. 

### Mutate

At this time, we don't know if we can/should create any new columns based on the data in the existing structure. 

```{r}
TelcoDataRaw %>%
  count(InternetService)
```


